IALE 2015 Analysis

### 2015-07-03 iale2015_sem.Rmd
Changed the calculation of understory richness by limiting the number of different species. Moved where I did the join of richness and UMCA data. Created an oak infection data frame joining UMCA infection, richness, and environmental variables. Created a pairs plot of this data frame.

veg <- rbind(veg_us_2005, veg_us_2011)
veg
unique(veg$year)
veg$year[veg$year < 2011] <- 2005 # Creates the two most complete veg sampling
veg <- veg %>% rename(us_species = UnderstorySp, plotid = PlotID)
richness <- veg %>% group_by(plotid, year) %>% summarise(us_rich = length(unique(us_species)))
richness
```

@@

stems <- as.tbl(stems)
summary(stems)
unique(stems$status)
stems <- stems %>% select(-X) %>% filter(status == "Alive" | status == "Dead")
unique(stems$species)
stems$species <- tolower(stems$species)
stems$species[stems$species == "quke x quag"] <- "quag x quke"
stems$plotid <- tolower(stems$plotid)
stems %>% group_by(plotid, year) %>% summarise(tagged_rich = length(unique(species)))

os_richness <- left_join(
      stems %>% group_by(plotid, year) %>% summarise(tagged_rich = length(unique(species))),
      untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species))), by = c("plotid", "year")
) %>% mutate(os_rich = tagged_rich + untagged_rich)
os_richness
richness <- left_join(os_richness, richness, by = c("plotid","year"))
richness
summary(richness)
```

@@

stems_dbh <- stems_dbh %>% mutate(rel_dens = live_count / all_live_count, rel_dom = live_tot_dbh / all_live_dbh, imp_value = 0.5*(rel_dens + rel_dom))
summary(stems_dbh)
```

plots_umca <- ungroup(plots_umca) %>% select(-starts_with("psi"), -plot, -year)

pairs(select(plots_umca, -plot, -year, -starts_with("psi")), lower.panel = panel.cor, diag.panel = panel.hist)
```

>>>>

veg <- rbind(veg_us_2005, veg_us_2011)

unique(veg$year)
veg$year[veg$year < 2011] <- 2005 # Creates the two most complete veg sampling
unique(veg$us_species)
veg <- veg %>% rename(us_species = UnderstorySp, plotid = PlotID)

# I want to match the species to those used by Sarah in her second chapter
local_woody <- read.csv("analysis/data/presence_local_woody_n197.csv")
summary(local_woody)
unique(local_woody$spp.local)
woody_spp <- as.character(local_woody$spp.local)
woody_spp[woody_spp == "prunus"] <- "prunus sp."
sort(woody_spp)
sort(unique(veg$us_species))
# It looks like I will want to add a few more to her list

filter(veg, us_species == "ribes sp.")
filter(veg, us_species == "ribes")
veg$us_species[veg$us_species == "ribes"] <- "ribes sp."

woody_spp <- c(woody_spp, "ceanothus sp.", "cebe", "cecu", "frla", "garrya sp.", "gemo", "juca", "mafu", "mane", "mimulus sp", "pisa", "quag x quke", "qube", "quercus sp.", "quwi", "rhamnus sp.", "rhcr", "ribes sp.", "rica", "risa", "roca", "rosa sp.", "rubus sp.", "rudi", "salix sp.", "syal", "vaccinium sp.")

veg_sub <- filter(veg, veg$us_species %in% woody_spp)

richness <- veg_sub %>%
      group_by(plotid, year) %>% 
      summarise(us_rich = length(unique(us_species)))
richness
```

@@

stems <- as.tbl(stems)
summary(stems)
unique(stems$status)

stems <- stems %>% select(-X) %>% filter(status == "Alive" | status == "Dead")
unique(stems$species)
stems$species <- tolower(stems$species)
stems$species[stems$species == "quke x quag"] <- "quag x quke"
stems$plotid <- tolower(stems$plotid)
stems %>% group_by(plotid, year) %>% summarise(tagged_rich = length(unique(species)))

os_richness <- left_join(
      stems %>% group_by(plotid, year) %>% summarise(tagged_rich = length(unique(species))),
      untagged_dbh %>% group_by(plotid, year) %>% summarise(untagged_rich = length(unique(species))), by = c("plotid", "year")
) %>% mutate(os_rich = tagged_rich + untagged_rich)
os_richness

richness <- left_join(os_richness, richness, by = c("plotid","year"))
richness
summary(richness)
```

Now I want to join the richness data set to the `plots_umca` data frame that currently has the environmental and leaf count/infection variables.

```{r join richness data}
plots_umca
richness
plots_umca <- rename(plots_umca, plotid = plot)
plots_umca$plotid <- tolower(plots_umca$plotid)

plots_umca <- left_join(plots_umca, richness, by = c("plotid", "year"))
```

plots_umca <- ungroup(plots_umca)

pairs(select(plots_umca, -plotid, -year, -starts_with("psi"), -tagged_rich, -untagged_rich), lower.panel = panel.cor, diag.panel = panel.hist)
```

Now that I have summarized some vegetation (not species specific), environmental, and bay laurel disease data to the plot-level I want to join this to the plot-level oak infection data.

```{r join umca and qusp data}
plots_umca
plot_qusp # right now this is broken down by oak species
summary(plot_qusp)
oak_sod <- plot_qusp %>% group_by(plot, year) %>%
      summarise(uninf_oak_ct = sum(uninfected_oak_ct), inf_oak_ct = sum(infected_oak_ct), tot_oak_ct = sum(c(uninfected_oak_ct, infected_oak_ct)))
oak_sod
oak_sod <- rename(oak_sod, plotid = plot)
oak_sod$plotid <- tolower(oak_sod$plotid)

oak_sod <- left_join(oak_sod, plots_umca[1:7], 
                     by = c("plotid", "year"))
summary(oak_sod) # the NA's for the bay laurel count variables are legitimately zeroes b/c there are no bay in some of these plots
oak_sod$uninfected_bay_ct[is.na(oak_sod$uninfected_bay_ct)] <- 0
oak_sod$infected_bay_ct[is.na(oak_sod$infected_bay_ct)] <- 0
oak_sod$tot_bay[is.na(oak_sod$tot_bay)] <- 0

head(plots_env)
plots_env$plotid <- tolower(plots_env$plotid)
oak_sod <- left_join(oak_sod, plots_env, by = c("plotid", "year" = "sample_year"))
oak_sod

richness
oak_sod <- left_join(oak_sod, richness, by = c("plotid", "year"))
summary(oak_sod)
summary(filter(oak_sod, is.na(avg_tmax)))
filter(oak_sod, is.na(avg_tmax), year == 2005) # BUSH01 plot; I'm dropping that from the data set
oak_sod <- filter(oak_sod, plotid != "bush01")
summary(filter(oak_sod, is.na(avg_tmax)))
filter(oak_sod, is.na(avg_tmax), year == 2004) # PONTI01 and SWEET01 were other early abandoned plots
oak_sod <- filter(oak_sod, plotid != "ponti01", plotid != "sweet01")
summary(filter(oak_sod, is.na(avg_tmax))) # Now the NAs in the environmental variables are just for year 2003 when we had severely incomplete temperature data
summary(oak_sod)

filter(oak_sod, plotid == "ann05") # There is no 2007 record for ann05 in the UMCA form, but there is a plot visit. I will need to come back and investigate this further for correcting the database.
oak_sod$infected_bay_ct[oak_sod$plotid == "ann05" & oak_sod$year == 2007] <- 15
oak_sod$tot_bay[oak_sod$plotid == "ann05" & oak_sod$year == 2007] <- 15
# I filled in the counts based on the values in the immediately preceding and following years, which I think is pretty solid reasoning. However, there is nothing I can do about the leaf counts right now.
summary(oak_sod)

filter(oak_sod, year == 2003) # 87 records
filter(oak_sod, year == 2004) # 90 records
# 2005 appears to be the first full year of observing all the plots
filter(oak_sod, year == 2005) # 171 records
filter(oak_sod, year == 2006) # 172 records
filter(oak_sod, year == 2007) # 171 records
filter(oak_sod, year == 2008) # 171 records
filter(oak_sod, year == 2009) # 170 records
filter(oak_sod, year == 2010) # 170 records
filter(oak_sod, year == 2011) # 170 records; final year of Sarah's 2nd chapter
filter(oak_sod, year == 2012) # 169 records
filter(oak_sod, year == 2014) # 168 records
# In Sarah's analysis we end up using only 163 plots
```

Now to do a pairs plot of the `oak_sod` data frame

```{r oak_sod pairs plot}
class(oak_sod)
oak_sod <- ungroup(oak_sod)

pairs(select(oak_sod, -plotid, -year, -starts_with("psi"), -tagged_rich, -untagged_rich), lower.panel = panel.cor, diag.panel = panel.hist)
```

@@

stems_dbh <- stems_dbh %>% mutate(rel_dens = live_count / all_live_count, rel_dom = live_tot_dbh / all_live_dbh, imp_value = 0.5*(rel_dens + rel_dom))
summary(stems_dbh)
write.csv(stems_dbh, "analysis/data/species_dbh_impval_plot.csv")
```
